### 2024-05-31 journal/correctness.tex
working on reviewer 2's comments

If $\var{v}$ is a leaf, $B$ was created at line \ref{enqNew} or \ref{deqNew}.
For line \ref{enqNew}, $B$ represents a single enqueue, so $|E(B)|=1$ and $|D(B)|=0$.
Since $B.\fld{sum\sub{enq}}$ is set to $\var{v}.\fld{blocks}[i-1].\fld{sum\sub{enq}}+1$ and
$B.\fld{sum\sub{deq}}$ is set to $\var{v}.\fld{blocks}[i-1].\fld{sum\sub{deq}}$, the claim follows from the hypothesis.
The proof for line~\ref{deqNew}, where $B$ represents a single dequeue, is similar.

Now suppose $\var{v}$ is an internal node. By the definition of \rev{direct} subblocks in (\ref{defsubblock}) and Lemma \ref{lem::headProgress}, the \rev{direct}
subblocks of $\var{v}.\fld{blocks}[1..i]$ are $\var{v}.\fld{left.blocks}[1..B.\eleft]$ 
and $\var{v}.\fld{right.blocks}[1..B.\eright]$.
Thus, the enqueues in $E(\var{v}.\fld{blocks}[0])\cdots E(\var{v}.\fld{blocks}[i])$ are those in
$E(\var{v}.\fld{left.blocks}[0]) \cdots E(\var{v}.\fld{left.blocks}[B.\eleft])$ and those in 
$E(\var{v}.\fld{left.blocks}[0]) \cdots E(\var{v}.\fld{left.blocks}[B.\eright])$.
By the hypothesis, the total number of these enqueues is 
$\var{v}.\fld{left.blocks}[B.\eleft].\fld{sum\sub{enq}} + \var{v}.\fld{right.blocks}[B.\eright].\fld{sum\sub{enq}}$, 
which is the value that line \ref{createSumEnq} stored in $B.\fld{sum\sub{enq}}$ when $B$ was created.
The proof for \fld{sum\sub{deq}} (stored on line~\ref{createSumDeq}) is similar.
\end{proof}

@@

%\Eric{I think this case was missing from the proof in the thesis}
Now suppose $R$'s call to \op{CreateBlock} returns \nl, causing $R$ to terminate at line \ref{addOP}.
Intuitively, since there are no operations in $\var{v}$'s children to promote, $op$ is already in $\var{v}$.
We formalize this intuition.
The value computed at line \ref{computeNumEnq} is
\begin{eqnarray*}
\var{num\sub{enq}}
&=&  \var{v}.\fld{left.blocks}[B.\eleft].\fld{sum\sub{enq}} \\
&& + \var{v}.\fld{right.blocks}[B.\eright].\fld{sum\sub{enq}}\\
&&  - \var{v}.\fld{blocks}[h-1].\fld{sum\sub{enq}} \\
&=& \var{v}.\fld{left.blocks}[B.\eleft].\fld{sum\sub{enq}}\\
&& + \var{v}.\fld{right.blocks}[B.\eright].\fld{sum\sub{enq}} \\
&&- \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[h-1].\eleft].\fld{sum\sub{enq}} \\
&& - \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[h-1].\eright].\fld{sum\sub{enq}}.
\end{eqnarray*}
It follows from Invariant \ref{inv::sum} that $num\sub{enq}$ is the total  number of enqueues  in 
$\var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[h-1].\eleft+1..B.\eleft]$ and
$\var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[h-1].\eright+1..B.\eright]$.
Similarly, $num\sub{deq}$ is the total number of dequeues contained in these blocks.
Since $num\sub{enq}+num\sub{deq}=0$ at line \ref{testEmpty},
these blocks contain no operations.
By Corollary \ref{blockNotEmpty}, this means the ranges of blocks are empty, so that $\var{v}.\fld{blocks}[h-1].\eleft \geq \var{B}.\eleft \geq i$.
Hence, $B_\ell$ is already a subblock of some block in $\var{v}$, so $\var{v}$ contains $op$.
\end{proof}

@@

\begin{lemma} \label{lem::appendExactlyOnce}
When an \op{Append}($B$) terminates, $B$'s operation is contained in exactly one block in each node along the path from the process's leaf to the root.
\end{lemma}
\begin{proof}
\op{Append} adds $B$ to the process's leaf and calls \op{Propagate}, which
does a double \op{Refresh}~on each internal node on the path $P$ from the leaf to the root.
By Lemma \ref{lem::doubleRefresh}, this ensures a block in each node on $P$ contains $B$'s operation.
There is at most one such block in each node, by Corollary \ref{cor::noDuplicates}.
\end{proof}

@@

\begin{lemma}\label{lem::indexDequeue}
If $\var{v}.\fld{blocks}[b]$ has been propagated to the root and $1\leq i\leq |D(\var{v}.\fld{blocks}[b])|$, 
 then \op{IndexDequeue}($\var{v}, b, i$) returns $\langle b',i' \rangle$ such that the \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$ is the $(i')$th dequeue of $D(\var{root}.\fld{blocks}[b'])$.
\end{lemma}

@@

Next, the position of the required dequeue in $D(B')$ (as defined by Equation (\ref{defSeqs})) is computed in 
lines \ref{computeISuperStart}--\ref{computeISuperEnd}. 
We first add the number of dequeues in the subblocks of $B'$ in $\var{v}$ that precede $B$ on line~\ref{computeISuperStart}.
If $\var{v}$ is the right child of its parent, then all of the subblocks of $B'$ from $\var{v}$'s left sibling
also precede the required dequeue, so we add the number of dequeues in those subblocks in line \ref{considerLeftBeforeRight}.

Finally, \op{IndexDequeue} is called recursively on $\var{v}$'s parent.
Since $B$ has been propagated to the root, so has its superblock $B'$.
Thus, all preconditions of the recursive call are met.
By the induction hypothesis, the recursive call returns the location of the required dequeue in the root.\end{proof}

\begin{lemma}\label{lem::get}
If $1\leq i\leq |E(\var{v}.\fld{blocks}[b])|$ then \op{getEnqueue}($\var{v},b,i$) returns the argument of the $i$th enqueue in $E(\var{v}.\fld{blocks}[b])$.
\end{lemma}

\begin{proof}
We prove the claim by induction on the height of node $\var{v}$.
If $\var{v}$ is a leaf, the hypothesis implies that $i=1$ and the block $\var{v}.\fld{blocks}[b]$ represents 
an enqueue whose argument is stored in $\var{v}.\fld{blocks}[b].\fld{element}$.
\op{GetEnqueue} returns the argument of this enqueue at line \ref{getBaseCase}.

Assuming the claim holds for $\var{v}$'s children, we prove it for $\var{v}$.
Let $B$ be $\var{v}.\fld{blocks}[b]$.
By Equation (\ref{defSeqs}),
$E(B)$ is obtained by concatenating the enqueue sequences of the direct subblocks
of $B$, which are listed in (\ref{defsubblock}).
By Invariant \ref{inv::sum}, $\var{sum\sub{left}}-\var{prev\sub{left}}$ is the number
of enqueues in $E(B)$ that come from $B$'s subblocks in $\var{v}$'s left child.
Thus, $dir$ is set to the direction for the child of $\var{v}$ that contains the required enqueue operation.
Moreover, when line \ref{endChooseDir} is reached, $i$ is the position of the required enqueue within the portion $E'$ of $E(B)$ that comes from that child.
Thus,  line \ref{getChild} finds the index $b'$ of the subblock $B'$ containing the required enqueue.
By Invariant \ref{inv::sum}, $\var{v}.\var{dir.blocks}[b'-1].\var{sum\sub{enq}} - \var{prev\sub{dir}}$ is the number of 
enqueues in $E'$ before the enqueues of block $B'$, so
the value $i'$ computed on line \ref{getChildIndex} is the position of the required enqueue within $E(B')$.
Thus, the recursive call on line \ref{getRecurse} satisfies its precondition, and 
returns the required result, by the induction hypothesis.
\end{proof}

@@

We show that the linearization ordering $L$ defined in Equation
(\ref{linearization}) is a legal permutation of a subset of the operations in 
the execution, i.e., that it includes all operations that terminate and 
if one operation $op_1$ terminates before another operation $op_2$ begins, then $op_2$ does not precede $op_1$ in $L$.  We also show each completed dequeue returns the same result as it would in the sequential execution~$L$.

@@

\begin{proof}
We prove the claim by induction on $b$. 
The base case when ${b=0}$ is trivially true, since the queue is initially empty and 
$\var{root}.\fld{blocks}[0]$ contains an empty block whose \size\ field is $0$. 
Assuming the claim holds for $b-1$, we prove it for $b$.
The \fld{size} field of the block $B$ installed in $\var{root}.\fld{blocks}[b]$ is computed
at line \ref{computeLength} of a call to \op{CreateBlock}(\var{root, b}).
By the induction hypothesis, $root.\fld{blocks}[b-1].\size$ gives the size of the queue before the operations
of block $B$ are performed.
By Invariant \ref{inv::sum}, the values of \var{num\sub{enq}} and \var{num\sub{deq}}
are the number of enqueues and dequeues contained in $B$.
Hence, the size of the queue after the operations of $B$ are performed (with enqueues before dequeues as specified by~$L$)
is $\max(0, \var{root}.\fld{blocks}[b-1].\size + \var{num\sub{enq}} - \var{num\sub{deq}})$.
\end{proof}

@@

\begin{lemma}\label{linearCorrect}
Each terminating dequeue returns the response it would in the sequential execution~$L$.
\end{lemma}
\begin{proof}
If a dequeue $Deq$ terminates, it is contained in some block in the root, by Lemma \ref{lem::appendExactlyOnce}.
By Lemma \ref{lem::indexDequeue}, $Deq$'s call to \op{IndexDequeue} on line~\ref{invokeIndexDequeue}
returns a pair $\langle b,i\rangle$ such that $Deq$ is the $i$th dequeue in the block 
$B=\var{root}.\fld{blocks}[b]$.
$Deq$ then calls \op{FindResponse}($b,i$) on line \ref{deqRest}.
By Lemma~\ref{sizeCorrectness}, the queue contains $\var{root}.\fld{blocks}[b-1].\fld{size}$ elements
after the operations in $\var{root}.\fld{blocks}[1..b-1]$ are performed sequentially 
in the order given by $L$.
By Invariant \ref{inv::sum}, the value of \var{num\sub{enq}} computed on line \ref{FRNum}
is the number of enqueues in $B$.
Since the enqueues in block $B$ precede the dequeues,
the queue is empty when the $i$th dequeue of $B$ occurs if 
$\var{root}.\fld{blocks}[b-1].\fld{size} + \var{num\sub{enq}} < i$.
So $Deq$ returns \nl\ on line \ref{returnNull} if and only if it would do so in the sequential
execution $L$.
Otherwise, the size of the queue after doing the operations in $\var{root}.\fld{blocks}[0..b-1]$
in the sequential execution $L$ is $\var{root}.\fld{blocks}[b-1].\fld{sum\sub{enq}}$ minus
the number of non-\nl\ dequeues in that prefix of $L$.
Hence, line \ref{computeE} sets $e$ to the rank of $Deq$ among all the non-\nl\ dequeues in $L$.
Thus, in the sequential execution~$L$, $Deq$ returns the value enqueued by the $e$th enqueue in $L$.
By Invariant \ref{inv::sum}, this enqueue is the $i_e$th enqueue 
in $E(\var{root}.\fld{blocks}[b_e])$, where
$b_e$ and $i_e$ are the values $Deq$ computes on line \ref{FRb} and \ref{FRi}.
By Lemma \ref{lem::get}, the call to \op{GetEnqueue} returns the argument of the required enqueue.
\end{proof}

>>>>

If $\var{v}$ is a leaf, $B$ was created at line \ref{enqNew} or \ref{deqNew}.
For line \ref{enqNew}, $B$ represents a single \enqueue, so $|E(B)|=1$ and $|D(B)|=0$.
Since $B.\fld{sum\sub{enq}}$ is set to $\var{v}.\fld{blocks}[i-1].\fld{sum\sub{enq}}+1$ and
$B.\fld{sum\sub{deq}}$ is set to $\var{v}.\fld{blocks}[i-1].\fld{sum\sub{deq}}$, the claim follows from the hypothesis.
The proof for line~\ref{deqNew}, where $B$ represents a single \dequeue, is similar.

Now, suppose $\var{v}$ is an internal node. By the definition of \rev{direct} subblocks in (\ref{defsubblock}) and Lemma \ref{lem::headProgress}, the \rev{direct}
subblocks of $\var{v}.\fld{blocks}[1..i]$ are $\var{v}.\fld{left.blocks}[1..B.\eleft]$ 
and $\var{v}.\fld{right.blocks}[1..B.\eright]$.
Thus, the \enqueues\ in $E(\var{v}.\fld{blocks}[0])\cdots E(\var{v}.\fld{blocks}[i])$ are those in
$E(\var{v}.\fld{left.blocks}[0]) \cdots E(\var{v}.\fld{left.blocks}[B.\eleft])$ and %those in 
$E(\var{v}.\fld{left.blocks}[0]) \cdots E(\var{v}.\fld{left.blocks}[B.\eright])$.
By the hypothesis, the total number of these \enqueues\ is 
$\var{v}.\fld{left.blocks}[B.\eleft].\fld{sum\sub{enq}} + \var{v}.\fld{right.blocks}[B.\eright].\fld{sum\sub{enq}}$, 
which is the value that line \ref{createSumEnq} stored in $B.\fld{sum\sub{enq}}$ when $B$ was created.
The proof for \fld{sum\sub{deq}} (stored on line~\ref{createSumDeq}) is similar.
\end{proof}

@@

%\Eric{I think this case was missing from the proof in the thesis}
Now suppose $R$'s call to \op{CreateBlock} returns \nl, causing $R$ to terminate at line \ref{addOP}.
Intuitively, since there are no operations in $\var{v}$'s children to promote, $op$ is already in $\var{v}$.
We formalize this intuition.
The value computed at line \ref{computeNumEnq} is
\begin{eqnarray*}
&&\var{num\sub{enq}}\\
&=&  \var{v}.\fld{left.blocks}[B.\eleft].\fld{sum\sub{enq}} \\
&& + \var{v}.\fld{right.blocks}[B.\eright].\fld{sum\sub{enq}}\\
&&  - \var{v}.\fld{blocks}[h-1].\fld{sum\sub{enq}} \\
&=& \var{v}.\fld{left.blocks}[B.\eleft].\fld{sum\sub{enq}}\\
&& + \var{v}.\fld{right.blocks}[B.\eright].\fld{sum\sub{enq}} \\
&&- \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[h-1].\eleft].\fld{sum\sub{enq}} \\
&& - \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[h-1].\eright].\fld{sum\sub{enq}}.
\end{eqnarray*}
It follows from Invariant \ref{inv::sum} that $num\sub{enq}$ is the total  number of \enqueues\  in 
$\var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[h-1].\eleft+1..B.\eleft]$ and
$\var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[h-1].\eright+1..B.\eright]$.
Similarly, $num\sub{deq}$ is the total number of \dequeues\ contained in these blocks.
Since $num\sub{enq}+num\sub{deq}=0$ at line \ref{testEmpty},
these blocks contain no operations.
By Corollary \ref{blockNotEmpty}, this means the ranges of blocks are empty, so that $\var{v}.\fld{blocks}[h-1].\eleft \geq \var{B}.\eleft \geq i$.
Hence, $B_\ell$ is already a subblock of some block in $\var{v}$, so $\var{v}$ contains $op$.
\end{proof}

@@

\begin{lemma} \label{lem::appendExactlyOnce}
When an \op{Append}($B$) terminates, $B$'s operation is contained in exactly one block in each node along the path from the process's leaf to the root.
\end{lemma}
\begin{proof}
\op{Append} adds $B$ to the process's leaf and calls \op{Propagate}, which
does a double \op{Refresh}~on each internal node on the path $P$ from the leaf to the root.
By Lemma \ref{lem::doubleRefresh}, this ensures a block in each node on $P$ contains $B$'s operation.
There is at most one such block in each node, by Corollary~\ref{cor::noDuplicates}.
\end{proof}

@@

\begin{lemma}\label{lem::indexDequeue}
If $\var{v}.\fld{blocks}[b]$ has been propagated to the root and $1\leq i\leq |D(\var{v}.\fld{blocks}[b])|$, 
 then \op{IndexDequeue}($\var{v}, b, i$) returns $\langle b',i' \rangle$ such that the \var{i}th \dequeue\ in $D(\var{v}.\fld{blocks}[\var{b}])$ is the $(i')$th \dequeue\ of $D(\var{root}.\fld{blocks}[b'])$.
\end{lemma}

@@

Next, the position of the required \dequeue\ in $D(B')$ (as defined by Equation (\ref{defSeqs})) is computed in 
lines \ref{computeISuperStart}--\ref{computeISuperEnd}. 
We first add the number of \dequeues\ in the subblocks of $B'$ in $\var{v}$ that precede $B$ on line~\ref{computeISuperStart}.
If $\var{v}$ is the right child of its parent, then all of the subblocks of $B'$ from $\var{v}$'s left sibling
also precede the required \dequeue, so we add the number of \dequeues\ in those subblocks in line \ref{considerLeftBeforeRight}.

Finally, \op{IndexDequeue} is called recursively on $\var{v}$'s parent.
Since $B$ has been propagated to the root, so has its superblock $B'$.
Thus, all preconditions of the recursive call are met.
By the induction hypothesis, the recursive call returns the location of the required \dequeue\ in the root.\end{proof}

\begin{lemma}\label{lem::get}
If $1\leq i\leq |E(\var{v}.\fld{blocks}[b])|$ then \op{getEnqueue}($\var{v},b,i$) returns the argument of the $i$th \enqueue\ in $E(\var{v}.\fld{blocks}[b])$.
\end{lemma}

\begin{proof}
We prove the claim by induction on the height of node $\var{v}$.
If $\var{v}$ is a leaf, the hypothesis implies that $i=1$ and the block $\var{v}.\fld{blocks}[b]$ represents 
an \enqueue\ whose argument is stored in $\var{v}.\fld{blocks}[b].\fld{element}$.
\op{GetEnqueue} returns the argument of this \enqueue\ at line \ref{getBaseCase}.

Assuming the claim holds for $\var{v}$'s children, we prove it for $\var{v}$.
Let $B$ be $\var{v}.\fld{blocks}[b]$.
By Equation (\ref{defSeqs}),
$E(B)$ is obtained by concatenating the \enqueue\ sequences of the direct subblocks
of $B$, which are listed in (\ref{defsubblock}).
By Invariant \ref{inv::sum}, $\var{sum\sub{left}}-\var{prev\sub{left}}$ is the number
of \enqueues\ in $E(B)$ that come from $B$'s subblocks in $\var{v}$'s left child.
Thus, $dir$ is set to the direction for the child of $\var{v}$ that contains the required \enqueue\ operation.
Moreover, when line \ref{endChooseDir} is reached, $i$ is the position of the required \enqueue\ within the portion $E'$ of $E(B)$ that comes from that child.
Thus,  line \ref{getChild} finds the index $b'$ of the subblock $B'$ containing the required \enqueue.
By Invariant \ref{inv::sum}, $\var{v}.\var{dir.blocks}[b'-1].\var{sum\sub{enq}} - \var{prev\sub{dir}}$ is the number of 
\enqueues\ in $E'$ before the \enqueues of block $B'$, so
the value $i'$ computed on line \ref{getChildIndex} is the position of the required \enqueue\ within $E(B')$.
Thus, the recursive call on line \ref{getRecurse} satisfies its precondition, and 
returns the required result, by the induction hypothesis.
\end{proof}

@@

We show that the linearization ordering $L$ defined in Equation
(\ref{linearization}) is a legal permutation of a subset of the operations in 
the concurrent execution, i.e., that it includes all operations that terminate and 
if one operation $op_1$ terminates before another operation $op_2$ begins, then $op_2$ does not precede $op_1$ in $L$.  We also show each completed \dequeue\ returns the same result \rev{in the concurrent execution} as it would in the sequential execution~$L$ \rev{(i.e., the sequential execution of the operations in the order given by $L$)}.

@@

\begin{proof}
We prove the claim by induction on $b$. 
The base case when ${b=0}$ is trivially true, since the queue is initially empty and 
$\var{root}.\fld{blocks}[0]$ contains an empty block whose \size\ field is $0$. 
Assuming the claim holds for $b-1$, we prove it for $b$.
The \fld{size} field of the block $B$ installed in $\var{root}.\fld{blocks}[b]$ is computed
at line \ref{computeLength} of a call to \op{CreateBlock}(\var{root, b}).
By the induction hypothesis, $root.\fld{blocks}[b-1].\size$ gives the size of the queue before the operations
of block $B$ are performed.
By Invariant \ref{inv::sum}, the values of \var{num\sub{enq}} and \var{num\sub{deq}}
are the number of \enqueues\ and \dequeues\ contained in $B$.
Hence, the size of the queue after the operations of $B$ are performed (with \enqueues\ before \dequeues\ as specified by~$L$)
is $\max(0, \var{root}.\fld{blocks}[b-1].\size + \var{num\sub{enq}} - \var{num\sub{deq}})$.
\end{proof}

@@

\begin{lemma}\label{linearCorrect}
Each terminating \dequeue\ returns the response it would in the sequential execution~$L$.
\end{lemma}
\begin{proof}
If a \dequeue\ $Deq$ terminates, it is contained in some block in the root, by Lemma \ref{lem::appendExactlyOnce}.
By Lemma \ref{lem::indexDequeue}, $Deq$'s call to \op{IndexDequeue} on line~\ref{invokeIndexDequeue}
returns a pair $\langle b,i\rangle$ such that $Deq$ is the $i$th \dequeue\ in the block 
$B=\var{root}.\fld{blocks}[b]$.
$Deq$ then calls \op{FindResponse}($b,i$) on line \ref{deqRest}.
By Lemma~\ref{sizeCorrectness}, the queue contains $\var{root}.\fld{blocks}[b-1].\fld{size}$ elements
after the operations in $\var{root}.\fld{blocks}[1..b-1]$ are performed sequentially 
in the order given by $L$.
By Invariant \ref{inv::sum}, the value of \var{num\sub{enq}} computed on line \ref{FRNum}
is the number of \enqueues\ in $B$.
Since the \enqueues\ in block $B$ precede the \dequeues,
the queue is empty when the $i$th \dequeue\ of $B$ occurs if 
$\var{root}.\fld{blocks}[b-1].\fld{size} + \var{num\sub{enq}} < i$.
So $Deq$ returns \nl\ on line \ref{returnNull} if and only if it would do so in the sequential
execution $L$.
Otherwise, the size of the queue after doing the operations in $\var{root}.\fld{blocks}[0..b-1]$
in the sequential execution $L$ is $\var{root}.\fld{blocks}[b-1].\fld{sum\sub{enq}}$ minus
the number of non-\nl\ \dequeues\ in that prefix of $L$.
Hence, line \ref{computeE} sets $e$ to the rank of $Deq$ among all the non-\nl\ \dequeues\ in $L$.
Thus, in the sequential execution~$L$, $Deq$ returns the value enqueued by the $e$th \enqueue\ in $L$.
By Invariant \ref{inv::sum}, this \enqueue\ is the $i_e$th \enqueue\ in $E(\var{root}.\fld{blocks}[b_e])$, where
$b_e$ and $i_e$ are the values $Deq$ computes on line \ref{FRb} and \ref{FRi}.
By Lemma \ref{lem::get}, the call to \op{GetEnqueue} returns the argument of the required \enqueue.
\end{proof}

### 2024-05-31 journal/introduction.tex
working on reviewer 2's comments

The lock-free \emph{MS-queue} of Michael and Scott \cite{MS98} is a classic shared queue implementation.
It uses a singly-linked list with pointers to the front and back nodes.
To dequeue or enqueue an element, the front or back pointer is updated by a 
compare-and-swap (CAS) instruction.
If this CAS fails, the operation must retry.
In the worst case, this means that each successful CAS may cause all other processes to
fail and retry, leading to an amortized step complexity of $\Omega(p)$ per operation in a system of $p$ processes.
Numerous papers have suggested modifications to the MS-queue \cite{DBLP:conf/opodis/HoffmanSS07,DBLP:conf/podc/KoganH14,DBLP:conf/ppopp/KoganP11,DBLP:journals/dc/Ladan-MozesS08,MKLLP22,DBLP:conf/spaa/MoirNSS05,RC17}, but 
all still have $\Omega(p)$ amortized step complexity as a result of
contention on the front and back of the queue.
Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} called this the \emph{CAS retry problem}.
The same problem occurs in array-based implementations of queues \cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/opodis/GidenstamST10,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01}.
Solutions that tried to sidestep this problem using fetch\&increment \cite{DBLP:conf/ppopp/MorrisonA13,Nik19,10.1145/3490148.3538572,DBLP:conf/ppopp/YangM16}
rely on slower mechanisms to handle worst-case executions and still have $\Omega(p)$ step complexity.

@@

Our queue uses a binary tree, called the \emph{\ordering\ tree}, where each process has its own leaf.
A process adds its operations to its leaf.
As in previous work (e.g., \cite{DBLP:conf/stoc/AfekDT95,DBLP:conf/fsttcs/JayantiP05}), operations are propagated from the leaves up to the root in a cooperative way that ensures wait-freedom
and avoids the CAS retry problem.
Operations reach the root in batches and are linearized in the order they reach the root.
Since a batch can have up to $p$ operations, explicitly recording the list of operations composing a batch
or applying them one-by-one to the queue would be too costly.
Instead, we use a novel implicit representation of batches
of queue operations that allows us to quickly merge two batches from the children of a node,
and quickly access any  operation in a~batch.
Preliminary versions of this work appeared in~\cite{Nad22,NR23}.
%\here{Maybe say a little more about techniques used in the implementation, if space permits}

>>>>

The lock-free \emph{MS-queue} of Michael and Scott \cite{MS98} is a classic shared queue implementation.
It uses a singly-linked list with pointers to the front and back nodes.
To dequeue or enqueue an element, the front or back pointer is updated by a 
compare-and-swap (CAS) instruction\footnote{\rev{A CAS(\var{X}, \var{old}, \var{new}) instruction updates a variable \var{X} to the value \var{new} and returns \tr\ if \var{X}'s value is \var{old}; otherwise the CAS leaves \var{X} unchanged and returns \fa.}}.
If this CAS fails, the operation must retry.
In the worst case, this means that each successful CAS may cause all other processes to
fail and retry, leading to an amortized step complexity of $\Omega(p)$ per operation in a system of $p$ processes.
Numerous papers have suggested modifications to the MS-queue \cite{DBLP:conf/opodis/HoffmanSS07,DBLP:conf/podc/KoganH14,DBLP:conf/ppopp/KoganP11,DBLP:journals/dc/Ladan-MozesS08,MKLLP22,DBLP:conf/spaa/MoirNSS05,RC17}, but 
all still have $\Omega(p)$ amortized step complexity as a result of
contention on the front and back of the queue.
Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} called this the \emph{CAS retry problem}.
The same problem occurs in array-based implementations of queues \cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/opodis/GidenstamST10,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01}.
Solutions that tried to sidestep this problem using fetch\&increment \cite{DBLP:conf/ppopp/MorrisonA13,Nik19,10.1145/3490148.3538572,DBLP:conf/ppopp/YangM16}
rely on slower mechanisms to handle worst-case executions and still have $\Omega(p)$ step complexity.

@@

Our queue uses a binary tree, called the \emph{\ordering\ tree}, where each process has its own leaf.
A process adds its operations to its leaf.
As in previous work (e.g., \cite{DBLP:conf/stoc/AfekDT95,DBLP:conf/fsttcs/JayantiP05}), operations are propagated from the leaves up to the root in a cooperative way that ensures wait-freedom
and avoids the CAS retry problem.
Operations reach the root in batches and are linearized in the order they reach the root.
Since a batch can have up to $p$ operations, explicitly recording the list of operations composing a batch
or applying them one-by-one to the queue would be too costly.
Instead, we use a novel implicit representation of batches
of queue operations that allows us to quickly merge two batches from the children of a node,
and quickly access any  operation in a~batch.
Preliminary \rev{versions of this work appeared in~\cite{Nad22,NR23}}.
%\here{Maybe say a little more about techniques used in the implementation, if space permits}
